---
title: "Group_assignment_clean"
author: "Tom Kugener"
date: "3/26/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction

# Methods 
# Results 
# Discussion 

# Complete R code
# Before getting started
```{r, warning=FALSE, message=FALSE}
# Install and load necessary packages 
source("Functions.R")
list.of.packages <- c("haven", "dplyr", "reshape","reshape2", "magrittr", "httr","kableExtra", "ggplot2", "GGally", "epiDisplay", "tidyverse", "caret", "leaps", "arsenal", "skimr") 
installRequiredPackages(list.of.packages)

# Import Data
pathToData <- '.' 
pancreatitis <- read.csv( file.path(pathToData,'pancreatitis.csv') )
```


# Recoding & Descriptive statistics 
```{r}
# To allow summary statistics to work, the level of measurement of the variables will be defined
recoded_pancreatitis <- dplyr::mutate(pancreatitis,
                                      site = as.factor(site),
                                      age = as.numeric(age),
                                      risk = as.numeric(risk),
                                      gender = as.factor(gender),
                                      outcome = as.factor(outcome),
                                      sod = as.factor(sod),
                                      pep = as.factor(pep),
                                      recpanc = as.factor(recpanc) ,
                                      psphinc = as.factor(psphinc),
                                      precut = as.factor(precut),
                                      difcan = as.factor(difcan), 
                                      pneudil = as.factor(pneudil),
                                      amp = as.factor(amp),
                                      paninj = as.factor(paninj),
                                      acinar = as.factor(acinar),
                                      brush = as.factor(brush),
                                      asa81 = as.factor(asa81),
                                      asa325 = as.factor(asa325),
                                      asa = as.factor(asa),
                                      prophystent = as.factor(prophystent), 
                                      therastent = as.factor(therastent),  
                                      pdstent = as.factor(pdstent),       
                                      sodsom = as.factor(sodsom),
                                      bsphinc = as.factor(bsphinc),
                                      bstent = as.factor(bstent), 
                                      chole = as.factor(chole),
                                      pbmal = as.factor(pbmal),
                                      train = as.factor(train), 
                                      status = as.factor(status),
                                      type = as.factor(type),
                                      rx = as.factor(rx),
                                      bleed = as.factor(bleed)      
                                      )

summary(recoded_pancreatitis, text = TRUE, title = "Descriptive Statistics on Pancreatitis")
library(arsenal) 
require(knitr)
require(survival)
t3 <- tableby(~ ., data = recoded_pancreatitis)
summary(t3, text = TRUE, title = "Descriptive Statistics on Pancreatitis")
summary(recoded_pancreatitis)
skim(recoded_pancreatitis)
```



# Selection of candidate variables
Based on established literature, and considering the  practical availability of certain variables, a subset of the variables was chosen as candidate predictors
```{r}
subset_pancreatitis <- pancreatitis[,names(pancreatitis) %in% c("rx", "age", "gender", "amp", "pep", "train", "chole", "difcan", "recpanc", "sod", "pdstent")]

subset_recoded_pancreatitis <- recoded_pancreatitis[,names(recoded_pancreatitis) %in% c("rx", "age", "gender", "amp", "pep", "train", "chole", "difcan", "recpanc", "sod", "pdstent")]

# Summary of candidate predictors
summary(subset_recoded_pancreatitis)

# Check normal distribution of numerical variable
hist(pancreatitis$age)
qqnorm(pancreatitis$age, pch = 1, frame = FALSE)
qqline(pancreatitis$age, col = "green", lwd = 1)

#Check linear association between age and outcome (is this relevant for a binary outcome?)
plot(recoded_pancreatitis$outcome, recoded_pancreatitis$age)

```
# Logistic Regression model & Backward stepwise regression
```{r}
# Backward regression

install.packages("MASS")
library(MASS)
library(rms)

model_p1 <-  glm(outcome ~ rx + age + gender + amp +
                   pep + train+ chole + difcan + recpanc + sod+ pdstent,
                 family = binomial, data = pancreatitis)
model_p1

model_rms_p1 <- lrm(data = pancreatitis, outcome ~ rx + age + gender + amp +
                      pep + train+ chole + difcan + recpanc + sod+ pdstent,
                    x = TRUE, y = TRUE)
model_rms_p1

step.model <- stepAIC(model_p1, direction = "backward", 
                      trace = FALSE)
summary(step.model)
?stepAIC



# LASSO
library(glmnet)
x <- data.matrix(pancreatitis[,c('rx', 'age', 'gender' ,'amp',
                      'pep', 'train', 'chole' ,'difcan' ,'recpanc' , 'sod', 'pdstent')])
y <- pancreatitis$outcome
lasso_model <- cv.glmnet(x,y, alpha = 1)
best_lambda <- lasso_model$lambda.min
best_lambda
plot(lasso_model)
coef(lasso_model)

best_model <- glmnet(x, y, alpha = 1, lambda = best_lambda) ## the best lambda is really small and therefore does not lead to the exclusion of variables
coef(best_model)

best_model <- glmnet(x, y, alpha = 1, lambda = 0.005)
coef(best_model)

best_model <- glmnet(x, y, alpha = 1, lambda = 0.01)
coef(best_model)
```
# Validation 
```{r}
model_pf <-  glm(outcome ~ rx + age +  amp +
                   pep + train+ chole + difcan + recpanc + sod,
                 family = binomial, data = pancreatitis)

model_pf

model_rms_pf <- lrm(data = pancreatitis, outcome ~ rx + age +  amp +
                      pep + train+ chole + difcan  + sod,
                    x = TRUE, y = TRUE)
model_rms_pf

validation_rms <- validate(model_rms_pf, method= "boot", B=500)
validation_rms

## Just checking whether recoding matters 
model_rms_pf2 <- lrm(data = recoded_pancreatitis, outcome ~ rx + age +  amp +
                      pep + train+ chole + difcan  + sod,
                    x = TRUE, y = TRUE)
model_rms_pf2

validation_rms2 <- validate(model_rms_pf2, method= "boot", B=500)
validation_rms2

plot(validation_rms, B=500)

# C statistic
cstatapp <- 0.5*(validation_rms[1,]+1)
cstatapp

library(rms)

# Calibration of the model
cal <- calibrate(model_rms_pf, B = 500)
plot(cal)
```


